Double Standards in Judging Collective Action\thanks{}

### 2024-04-16 Documents/manuscript.tex
resubmit

\noindent Collective action is a powerful force driving social change
but often sparks contention about what actions are acceptable means to
effect social change. In five studies (total \emph{N} = 2,979), we
investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In three
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police and for or against
restricting abortion in the United States. We found evidence for
ideology-based, but not for identity-based, double standards:
Participants judged the same protest actions to be more acceptable when
the protesters' cause aligned with their own ideological position---but
showed no consistent ingroup bias in terms of class, race, and gender.
Our findings have theoretical and practical implications for
understanding the often divided response to prominent social movements.

\textit{Keywords:} collective action, double standards, social identity
theory, system justification theory, Black Lives Matter

@@

\begin{center}
\noindent 11,546 words
\end{center}

@@

\noindent This research shows that what people consider an acceptable
\emph{means} of protest depends on what the \emph{end} of a protest is.
That is, people tend to judge the \emph{same} controversial protest
actions to be more acceptable when the protesters' cause aligned with
their own ideological positions. This finding helps explain the often
divided response to recent protest movements.

@@

Protest movements often spark contention about what actions are
acceptable means of protest. Reactions to Black American athletes
kneeling during the national anthem to protest racist police violence
are a case in point. Kneeling during the national anthem is not violent,
disruptive, or illegal. And yet, only 29\% of White Americans (compared
to 66\% of Black Americans) and 11\% of Republicans (compared to 59\% of
Democrats) considered it appropriate for Black athletes to protest in
this way (YouGov, 2017). Indeed, Black athletes who had participated in
the protests faced a backlash in public opinion (Lacina, 2020). This and
other examples suggest that \emph{who} the protesters are and
\emph{what} they are protesting influences how acceptable observers
judge their actions to be.

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the same protest actions as
more or less acceptable depending on their own and the protesters' group
memberships (\emph{identity-based double standards}) or on the
protesters' cause and how it aligns with their own ideological positions
(\emph{ideology-based double standards}). In two studies, we first
develop an instrument of 25 controversial protest actions, based in item
response theory, to capture double standards in judging collective
action. In three preregistered experiments, we then use this instrument
to test for double standards in judging protest actions for workers'
rights, for or against defunding the police, and for or against
restricting abortion. We find evidence for ideology-based, but not for
identity-based, double standards: People judge the same protest actions
to be more acceptable when the protesters' cause aligns with their own
ideological positions---but do not show consistent ingroup bias (in
terms of class, race, and gender) when judging collective action by
ingroup and outgroup members.

@@

Recent research has shown that this distinction matters for how people
respond to collective action. Shuman et al. (2021) showed that
non-normative, non-violent collective action is more effective than
either normative or violent collective action at gaining concessions
from those most resistant to social change. Feinberg et al. (2020)
demonstrated, however, that observers are less supportive of social
movements that use extreme protest actions.\footnote{In contrast, White
  Americans exposed to pictures of destructive Black Lives Matter
  protests were not, on average, any less supportive of systemic change
  than those exposed to pictures of peaceful protests (Teixeira et al.,
  2022).} Similarly, Teixeira et al. (2020) found that advantaged-group
members perceive non-normative collective action by disadvantaged-group
members as more damaging to their ingroup's social image and are thus
less supportive of such action. Verkuyten et al. (2022) found that, in
general, the public tends to be intolerant of non-normative collective
action. This means that activists face the dilemma that non-normative
collective action might be most effective at gaining concessions but
that it also risks reducing popular support for a cause. Together, these
studies show that the distinction between normative and non-normative
collective action is psychologically and societally consequential.

These and other studies have, for the most part, relied on ad-hoc
distinctions between what researchers themselves considered normative
and non-normative collective action. In liberal democracies, researchers
tend to consider actions such as signing petitions, voting in elections,
or peaceful protest to be normative and actions such as blocking
traffic, damaging property, or violent protest to be non-normative. But,
as reactions to Black athletes kneeling in protest suggest, the same
action might be considered acceptable by some and unacceptable by
others. That is, the distinction between normative and non-normative
collective action might be in the eye of the beholder.

In this research, we move this distinction from the researcher's
intuition into the realm of scientific investigation and test for double
standards in where people draw the line between acceptable (normative)
and unacceptable (non-normative) forms of collective action. For that
purpose, we define a double standard as judging the same protest actions
as more or less acceptable depending on who the protesters are and what
they are protesting (for an analogous definition, see Foschi,
2000).\footnote{Feinberg et al. (2020) provided incidental evidence for
  double standards in judging collective action as, in a manipulation
  check, Black participants rated Black Lives Matter protests as less
  extreme than White participants and liberal participants rated protest
  actions for progressive causes as less extreme than conservative
  participants.}

\hypertarget{identity-based-double-standards}{%
\subsubsection{Identity-Based Double
Standards}\label{identity-based-double-standards}}

@@

One domain in which ingroup bias manifests is judgments about other
people's actions. Hewstone (1990) reviewed research showing that people
make ingroup-serving causal attributions when judging actions by ingroup
and outgroup members, for example, by attributing negative behavior by
outgroup members to internal causes but attributing negative behavior by
ingroup members to external causes (see Pettigrew, 1979). Valdesolo \&
DeSteno (2007) demonstrated that participants judged the same selfish
action as less unfair when performed by themselves or an ingroup member
than when performed by an outgroup member. Other studies (Abrams et al.,
2013; Endevelt et al., 2021; for exceptions, see Mendoza et al., 2014;
Pinto et al., 2010) provided further evidence for ingroup bias in
judging moral transgressions by ingroup and outgroup members.

In the same vein, we propose that ingroup bias results in identity-based
double standards in judging collective action by ingroup and outgroup
members. That is, we hypothesize that people will judge the same protest
actions as more acceptable if the protesters are ingroup members
(\emph{Hypothesis 1a}) or if the cause of the protest aligns with the
ingroup's interests (\emph{Hypothesis 1b}). This hypothesis could
explain why, for example, far more Black than White Americans considered
it appropriate for Black athletes to kneel down to protest police
violence against Black Americans (YouGov, 2017). Notably, this
hypothesis applies equally to historically disadvantaged groups
mobilizing against social injustice and to historically advantaged
groups defending their group's position.

@@

System Justification Theory (Jost et al., 2004; for a recent review, see
Jost, 2020) argues that people are motivated to defend, justify, and
bolster the prevailing social, economic, and political system because
doing so serves basic epistemic, existential, and relational needs. Both
advantaged and disadvantaged groups are thought to be motivated to
justify the existing system, although the strength of this motivation
and its expression vary across individuals and situations.

@@

Just as system justification affects support for, and opposition to,
collective action, we propose that it results in ideology-based double
standards in what actions people consider acceptable forms of collective
action. First, we hypothesize that people will judge the same protest
actions as more acceptable if the protest supports, rather than
challenges, the system (\emph{Hypothesis 2a}). This hypothesis reflects
the assumption that the motivation to justify the system is, to some
extent, universal. Second, we hypothesize that people will judge the
same protest actions as more acceptable if they endorse
system-justifying beliefs and the protest supports the system or if they
reject system-justifying beliefs and the protest challenges the system
(\emph{Hypothesis 2b}). Notably, this hypothesis implies a symmetry in
how people with different ideological orientations judge
system-supporting and system-challenging collective action.

As system-justifying beliefs are closely related to political
conservatism (Jost et al., 2003), this hypothesis could explain partisan
differences in what actions liberals and conservatives consider
acceptable means of protest for progressive and conservative
causes---for example, why far more Democrats than Republicans considered
it appropriate to kneel during the national anthem to protest racist
police violence (YouGov, 2017).

\hypertarget{are-partisan-double-standards-based-in-ideology-or-identity}{%
\subsubsection{Are Partisan Double Standards Based in Ideology or
Identity?}\label{are-partisan-double-standards-based-in-ideology-or-identity}}

Partisan differences can result from two social psychological processes.
On the one hand, political partisanship reflects ideological differences
in resistance to social change and acceptance of inequality which, in
turn, are motivated by epistemic, existential, and relational needs
(Jost et al., 2003; Jost, 2021). As motivated cognition, ideological
differences shape how we think and feel about issues, actions, and
events. On the other hand, political partisanship functions as a social
identity (Finkel et al., 2020; Harris et al., 2022) with evidence
suggesting that partisans' policy preferences depend more on the stated
position of their party than policy content (Cohen, 2003). Like other
identities, political identities result in group-serving biases in how
we evaluate ingroup and outgroup members. In practice, the two processes
are often difficult to disentangle.

For now, we use ``identity-based'' to refer to double standards based on
assigned social categories such as class, race, and gender and
``ideology-based'' to refer to double standards based on ideological
alignment. We compare the evidence for identity and ideology processes
in the General Discussion.

@@

\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1}
\caption*{\textit{Note.} $\theta_{(z)}$ follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top).}
\label{fig:f1}
\end{figure}

In this research, item response theory serves two purposes. First, item
response theory provides a statistical framework for scale development.
In Studies 1 and 2, we select the most discriminating and informative
actions to build an instrument for capturing double standards in judging
collective action. Second, item response theory provides a formal and
statistical definition of double standards in judging collective action
as respondents being more accepting of the same protest actions
depending on who the protesters are and what they are protesting. In
Experiments 1--3, we operationalize double standards as the differences
in \(\theta_{(z)}\) between conditions which, as \(\theta_{(z)}\) is
\(z\)-standardized, correspond to Cohen's \(d\) effect sizes.

@@

The first purpose of our research is to develop an instrument to measure
where people draw the line between acceptable and unacceptable forms of
collective action. In Study 1, we ask participants to generate more and
less extreme protest actions and compile a pool of protest actions for
scale development. In Study 2, we ask participants to rate how
acceptable they consider each of those protest actions to be and use
item response theory to select the most discriminating and informative
protest actions for our instrument. In so doing, we build an instrument
that enables both present and future research to capture double
standards in judging collective action.

The second purpose of our research is to test for double standards in
judging collective action---that is, whether people judge the same
protest actions as more acceptable if the protesters' are ingroup
members or the protesters' cause aligns with the ingroup's interests
(\emph{Hypothesis 1a}, \emph{1b}) or if the protesters' cause aligns
with their own system-justifying or system-challenging motives
(\emph{Hypotheses 2a}, \emph{2b}).

In Experiment 1, we use the instrument developed in Studies 1 and 2 to
test for double standards in how acceptable participants consider
collective action for workers' rights in the United Kingdom. By varying
both the participants' and the protesters' group memberships
(working/middle class), Experiment 1 compares the relative evidence for
identity-based (\emph{Hypothesis 1a}, \emph{1b}) and ideology-based
double standards (\emph{Hypothesis 2a}). Extending Experiment 1,
Experiment 2 considers both system-challenging and system-defending
collective action in the context of the 2020 Black Lives Matter protests
in the United States. By varying the protesters' cause (for/against
defunding the police) as well as the participants' and the protesters'
group memberships (Black/White), Experiment 2 compares the relative
evidence for identity-based (\emph{Hypothesis 1a}, \emph{1b}) and
ideology-based double standards (\emph{Hypothesis 2a}, \emph{2b}).
Building on Experiments 1--2, Experiment 3 examines how liberal and
conservative men and women judge collective action for or against
restricting abortion in anticipation of the Supreme Court overturning
its \emph{Roe v. Wade} decision on reproductive rights in the United
States. By varying the protesters' cause as well as the participants'
ideology and gender, Experiment 3 compares the relative evidence for
identity-based (\emph{Hypothesis 1b}) and ideology-based
(\emph{Hypothesis 2b}) double standards in judging collective action
concerning a conservative cause in a sample balanced by gender and
ideology. Together, the preregistered experiments provide a
comprehensive test of the hypothesized double standards in judging
collective action.

@@

@@

\hypertarget{study-1}{%
\subsection{Study 1}\label{study-1}}

In Study 1, we compiled protest actions from participants' responses and
other sources. We recruited 60 participants from the Prolific subject
pool, all of whom were citizens of the UK or the US. To increase the
socioeconomic diversity of our sample, we recruited 30 non-students
without a university degree, 15 non-students with a university degree,
and 15 current university students.

Participants first read an accessible definition of what a social group
is and that collective action is any action group members take to
promote the interests of their social group. Participants were then
asked to name between five and ten actions that fit that definition.
Participants were encouraged to think of actions that varied in how
acceptable they were in their opinion. We recoded responses into a
smaller set of unique actions, which we supplemented with protest
actions from the psychological and political science literature (e.g.,
Sharp, 1973). This process resulted in 72 actions that varied in how
acceptable we would expect them to be. For details, see Supplemental
Online Material (SOM).

\hypertarget{study-2}{%
\subsection{Study 2}\label{study-2}}

@@

We estimated a graded response model (Bürkner, 2021; Samejima,
1997)---an item response theory model for ordinal response
variables---for participants' ratings of how often an action would be an
acceptable form of collective action.\footnote{Ratings of how often a
  given action would be an acceptable form of collective action were
  strongly and negatively correlated with how disruptive, violent, and
  extreme participants considered the action to be and how negative they
  felt about it (Table S3).} Based on the results, we selected the 25
most discriminating, informative, and relevant protest actions to form
an instrument for measuring double standards in judging collective
action in Experiments 1 and 2. We selected controversial, and thus
diagnostic, actions (e.g., disrupting traffic) over actions that almost
all respondents considered acceptable (e.g., signing petitions) or
unacceptable (e.g., political violence; Figure \ref{fig:f2}). For
detailed ratings and results, see SOM.

@@

In Experiment 1, we used the newly developed instrument to test for
double standards in judging collective action for workers' rights in the
United Kingdom. That is, we tested whether people with either
working-class or professional jobs applied different standards when
judging collective action taken by people with either working-class or
middle-class jobs to protest against a fictitious government bill
threatening their rights.

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that, in line with \emph{Hypothesis
1a} and \emph{1b}, participants would judge the same protest actions as
more acceptable if the protesters were ingroup members protesting for
the ingroup's rights than if they were outgroup members protesting for
the outgroup's rights. By comparing reactions to working-class and
middle-class protesters, we tested how observers judge collective action
by lower-status (i.e.~working-class) and higher-status (middle-class)
group members. Assuming that collective action by lower-status group
members is seen as more threatening to the unequal status quo than
collective action by higher-status group members and that people are, to
same extent, motivated to justify the system, we tested the
preregistered hypothesis that, in line with \emph{Hypothesis 2a},
participants would judge the same protest actions as more acceptable if
the protesters were from a higher-status group than if they were from a
lower-status group. In non-preregistered analyses, we further explored
whether participants would judge the same protest actions as more or
less acceptable depending on their own ideological positions. In this
way, Experiment 1 provided a first test of the hypothesized identity-
and ideology-based double standards in judging collective action.

@@

We recruited 515 participants from the Prolific subject pool who were UK
citizens, 25 years old or older, and not current students.\footnote{We
  had preregistered a sample of 500 participants, before exclusions, but
  included another 15 participants who completed the study without
  returning an approval code to the Prolific platform. Participants
  received, on average, \pounds 9.00 (\$10.93) per hour of
  participation.} As preregistered, we excluded 71 participants who
failed an attention check. This resulted in a final sample of 443
participants (\(\textit{Mdn} = 41\) years, age range: 25--76 years; 272
women, 171 men, 1 non-binary). Of these, 210 participants considered
their past, current, or future jobs to be working-class jobs.
Participants in this lower-status group did not have a university degree
and placed themselves on the bottom three ranks of the subjective
socioeconomic status ladder. Another 233 participants considered their
past, current, or future jobs to be middle-class/professional jobs.
Participants in this higher-status group had at least an undergraduate
degree and placed themselves on the top four ranks of the subjective
socioeconomic status ladder.

@@

Figure \ref{fig:f3} (A) shows estimates for each combination of the
protesters' and the participants' group membership. Overall, we found
that participants' responses depended on both the protesters' and the
participants' group status---but not in the directions predicted by our
hypotheses. Contradicting our first hypothesis, participants did not
consider protest actions performed by their ingroup to be more
acceptable, on average, than the same actions performed by the relevant
outgroup (Cohen's \(d = 0.00, [-0.20, 0.20]\)). Instead, we found that
participants from higher-status backgrounds considered protest actions
performed by both lower-status (\(d = 0.34, [0.07, 0.59\){]}) and
higher-status (\(d = 0.33, [0.03, 0.63]\)) protesters to be, on average,
more acceptable than participants from lower-status backgrounds.
Contradicting our second hypothesis, participants considered protest
actions performed by higher-status protesters to be less acceptable, on
average, than the same actions performed by lower-status protesters
(\(d = -0.20, [-0.40, -0.00]\)).

@@

Growing evidence shows that, in many situations, observers apply a more
lenient standard when judging moral transgressions by ingroup members
(Abrams et al., 2013; Endevelt et al., 2021). And yet, our findings
suggested that, unlike judgments in other domains, judgments about
collective action are not subject to double standards based on group
identities such as class, race, or gender. We did not find evidence for
those double standards even though we examined social categories that
are central to most people's identities, even when we considered only
highly identified group members, and even though participants' reactions
to the manipulation showed that, at least in Experiments 1--2, they
understood the protesters' causes in ingroup--outgroup terms.

>>>>

\noindent Collective action is a powerful force driving social change
but often sparks contention about what actions are acceptable means to
effect social change. Five studies (total \(N = 2,979\)) investigated
double standards in judging collective action---that is, whether
observers judge the same protest actions to be more acceptable depending
on who the protesters are and what they are protesting. Two studies used
item response theory to develop an instrument of 25 controversial
protest actions to measure where people draw the line between acceptable
and unacceptable forms of collective action. Three preregistered
experiments found no consistent evidence for ingroup bias in terms of
social class when judging protests for workers' rights (Experiment 1),
in terms of race when judging protests for and against defunding the
police (Experiment 2), and in terms of gender when judging protests for
and against restricting abortion (Experiment 3). Instead, we found that
participants who identified as left-wing/liberal (Experiments 1--3) and
rejected system-justifying beliefs (Experiments 1--2) considered the
same protest actions more acceptable when a cause aligned with their
ideological orientation (for workers' rights, for defunding the police,
against restricting abortion) than when it did not (against defunding
the police, for restricting abortion). Conservative participants
considered the same actions somewhat more acceptable when protesters
supported, rather than opposed, restricting abortion (Experiment 3) but
considered all protest actions, for and against defunding the police,
equally unacceptable (Experiment 2). Our findings have theoretical and
practical implications for understanding the often-divided response to
social movements.

\textit{Keywords:} collective action, double standards, social identity
theory, system justification theory, partisanship, Black Lives Matter,
abortion

@@

\begin{center}
\noindent 11,654 words
\end{center}

@@

\noindent This research shows that what people consider an acceptable
\emph{means} of protest depends on what the \emph{end} of a protest is.
That is, observers sometimes judge the \emph{same} controversial protest
action (e.g., blocking a road) to be more acceptable when the
protesters' cause (e.g., defunding the police) aligns with their own
ideological position. Conservatives, who support social order and oppose
social change, tend to consider all protest actions less acceptable than
progressives. Progressives, in turn, tend to consider disruptive but
non-violent actions acceptable means to achieve social change and
challenge social injustice. This finding helps explain the often-divided
response to recent protest movements.

@@

Protest movements often spark contention about what actions are
acceptable means of protest. Reactions to Black American athletes
kneeling during the national anthem to protest racist police violence
are a case in point. Kneeling during the national anthem is not violent,
disruptive, or illegal. And yet, only 29\% of White Americans (compared
to 66\% of Black Americans) and 11\% of Republicans (compared to 59\% of
Democrats) considered it appropriate for Black athletes to protest in
this way (YouGov, 2017). This and other examples suggest that \emph{who}
the protesters are and \emph{what} they are protesting influences how
acceptable their actions are judged to be.

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the \emph{same} protest
actions as more or less acceptable depending on their own and the
protesters' group memberships \emph{or} on the protesters' cause and how
it aligns with their own ideological positions.

@@

Recent research has shown that this distinction matters for how people
respond to collective action (for a review, see Shuman et al., 2023).
Shuman et al. (2021) showed that non-normative, non-violent collective
action is more effective than either normative or violent collective
action at gaining concessions from those most resistant to social
change. Feinberg et al. (2020) demonstrated, however, that observers are
less supportive of social movements that use extreme protest
actions.\footnote{In contrast, White Americans exposed to pictures of
  destructive Black Lives Matter protests were not, on average, any less
  supportive of systemic change than those exposed to pictures of
  peaceful protests (Teixeira et al., 2022).} Similarly, Teixeira et al.
(2020) found that advantaged-group members perceive non-normative
collective action by disadvantaged-group members as more damaging to
their ingroup's social image and are thus less supportive of such
action. This means that activists face the dilemma that non-normative
collective action might be most effective at gaining concessions but
that it also risks reducing popular support for a cause. Together, these
studies show that the distinction between normative and non-normative
collective action is psychologically and societally consequential.

Past research has, for the most part, relied on ad-hoc distinctions
between what researchers themselves considered normative and
non-normative collective action. In liberal democracies, researchers
tend to consider actions such as signing petitions, voting in elections,
or peaceful protest to be normative and actions such as blocking
traffic, damaging property, or violent protest to be non-normative. But,
as reactions to Black athletes kneeling in protest suggest, the same
action might be considered acceptable by some and unacceptable by
others. In this research, we test for double standards in where people
draw the line between what is an acceptable (normative) and an
unacceptable (non-normative) form of collective action.\footnote{In this
  research, we use \emph{acceptable} (``agreed or approved of by most
  people in a society,'' Oxford Advanced Learner's Dictionary, n.d.) as
  a plain-language synonym for \emph{normative} (``conforming to the
  norms of the existing social system'').} For that purpose, we define a
double standard as judging the same protest actions as more or less
acceptable depending on who the protesters are and what they are
protesting (for a similar definition, see Foschi, 2000).\footnote{Feinberg
  et al. (2020) provided incidental evidence for double standards in
  judging collective action as, in a manipulation check, Black
  participants rated Black Lives Matter protests as less extreme than
  White participants and liberal participants rated protest actions for
  progressive causes as less extreme than conservative participants.}

\hypertarget{group-based-double-standards}{%
\subsubsection{Group-Based Double
Standards}\label{group-based-double-standards}}

Far more Black (\(66\%\)) than White (\(29\%\)) Americans considered it
appropriate for Black athletes to kneel during the national anthem to
protest racist police violence (YouGov, 2017). This suggests that what
protest actions people consider acceptable might depend on their own and
the protesters' group memberships.

@@

One domain in which ingroup bias manifests is judgments about other
people's actions. Hewstone (1990) reviewed research showing that people
make ingroup-serving causal attributions when judging actions by ingroup
and outgroup members. For example, people tend to attribute negative
behavior by outgroup members to internal causes but attribute negative
behavior by ingroup members to external causes (see Pettigrew, 1979).
Valdesolo \& DeSteno (2007) demonstrated that participants judged the
same selfish action as less unfair when performed by themselves or an
ingroup member than when performed by an outgroup member. Other studies
(Abrams et al., 2013; Endevelt et al., 2021; for exceptions, see Mendoza
et al., 2014; Pinto et al., 2010) provided further evidence for ingroup
bias in judging moral transgressions by ingroup and outgroup members.

In the same vein, we propose that ingroup bias results in double
standards in judging collective action by ingroup and outgroup members.
That is, we hypothesize that people will judge the same protest actions
as more acceptable if the protesters are ingroup members and if the
cause of the protest aligns with the ingroup's interests. Notably, this
hypothesis applies equally to historically disadvantaged groups
mobilizing against social injustice and to historically advantaged
groups defending their group's position.

@@

Far more Democrats (\(59\%\)) than Republicans (\(11\%\)) considered it
appropriate to kneel during the national anthem to protest racist police
violence (YouGov, 2017). This suggests that what protest actions people
consider acceptable might depend on the protesters' cause and how it
aligns with their own ideological positions.

One explanation for ideology-based double standards is that they result
from the same intergroup processes as other group-based double
standards. In this view, political partisanship functions as a social
identity (Finkel et al., 2020; Harris et al., 2022) and results from
universal cognitive mechanisms for forming, detecting, and maintaining
alliances (Pinsof et al., 2023). Like other identities, political
identities result in group-serving biases in how we evaluate ingroup and
outgroup members. This explanation aligns with Verkuyten et al.'s (2022)
finding that participants were more tolerant of transgressive protest
actions (e.g., sending hate mail) when taken by their most-liked, rather
than their least-liked, political group (e.g., climate activists).

Another explanation for ideology-based double standards is that they are
rooted in ideological differences in how conservatives and progressives
relate to social change and social inequality which, in turn, are
motivated by epistemic, existential, and relational needs (Jost et al.,
2003; Jost, 2021). As motivated social cognition, ideological
differences shape how we think and feel about political issues, actions,
and events. In this view, ideology-based double standards depend on
whether the cause of a protest aligns with someone's ideological
positions.

System Justification Theory (Jost et al., 2004; for a recent review, see
Jost, 2020), in particular, argues that people are motivated to defend,
justify, and bolster the prevailing social, economic, and political
system because doing so serves basic epistemic, existential, and
relational needs. Both advantaged and disadvantaged groups are thought
to be motivated to justify the existing system, although the strength of
this motivation and its expression vary across individuals and
situations. Conservatives, far more so than progressives, endorse
system-justifying beliefs.

@@

Just as system justification affects support for, and opposition to,
collective action, we propose that it results in ideology-based double
standards in what actions people consider acceptable forms of collective
action. That is, we hypothesize that people will judge the same protest
actions as more acceptable if they endorse system-justifying beliefs and
the protest supports the system or if they reject system-justifying
beliefs and the protest challenges the system.

In principle, the two explanations for ideology-based double
standards---in terms of intergroup processes and ideological
differences---could be disentangled by separating political identities
and causes. An explanation in terms of intergroup processes posits that
political identities, not causes, shape partisan behavior. Indeed,
research suggests that partisans' policy preferences depend more on the
stated position of their party than policy content (Cohen, 2003). An
explanation in terms of ideological differences instead posits that
there is an ``elective affinity'' between political causes and the
psychological needs and motives of their supporters (Jost, 2021). In
other words, the content of political ideologies is not arbitrary or a
historical accident, as implied by an explanation in terms of intergroup
processes (see Pinsof et al., 2023). In practice, however, the two
explanations are difficult to disentangle in the context of real-world
protest movements in which political identities and causes can seldom be
separated.

For now, we use ``group-based'' to refer to double standards based on
ascribed categories such as class, race, and gender and
``ideology-based'' to refer to double standards based on ideological
alignment. We compare the evidence for different explanations of those
double standards in the General Discussion.

@@

\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1.png}
\caption*{\textit{Note.} $\theta_{(z)}$ follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top).}
\label{fig:f1}
\end{figure}

In this research, item response theory serves two purposes. First, it
provides a statistical framework for scale development. In Studies 1 and
2, we select the most discriminating and informative actions to build an
instrument for capturing double standards in judging collective action.
Second, it provides a formal and statistical definition of double
standards in judging collective action as respondents being more
accepting of the same protest actions depending on who the protesters
are and what they are protesting. In Experiments 1--3, we operationalize
double standards as differences in \(\theta_{(z)}\) which, as
\(\theta_{(z)}\) is \(z\)-standardized, correspond to Cohen's \(d\)
effect sizes.

@@

Five studies (total \(N = 2,979\)) investigate double standards in
judging collective actions---that is, whether observers judge the same
controversial protest actions to be more acceptable depending on who the
protesters are and what they are protesting.

Our first purpose is to develop an instrument to measure where people
draw the line between normative (acceptable) and non-normative
(unacceptable) forms of collective action. In Study 1, we ask
participants to generate more and less extreme protest actions and
compile a pool of protest actions for scale development. In Study 2, we
ask participants to rate how acceptable they consider each of those
protest actions to be and use item response theory to select the most
discriminating and informative protest actions for our instrument. In so
doing, we build an instrument that enables both present and future
research to capture double standards in judging collective action.

Our second purpose is to investigate potential double standards in
judging collective action. Three preregistered experiments used the
newly developed instrument to compare how acceptable participants
controversial protest actions in different conditions.

By varying the participants' and the protesters' group memberships, we
test for ingroup bias in terms of social class when judging protests for
workers' rights (Experiment 1), in terms of race when judging protests
for and against defunding the police (Experiment 2), and in terms of
gender when judging protests for and against restricting abortion
(Experiment 3). In this way, we test the hypothesis that people will
judge the same protest actions as more acceptable if the protesters are
ingroup members or the cause of the protest aligns with the ingroup's
interests.

By varying the protesters' causes, we test whether participants who
identify as left-wing/liberal (Experiments 1--3) and reject
system-justifying beliefs (Experiments 1--2) judge the same protest
actions to be more acceptable when the protesters' cause is progressive
and system challenging (for workers' rights, for defunding the police,
against restricting abortion). Likewise, we test whether participants
who identify as right-wing/conservative and endorse system-justifying
beliefs judge the same protest actions to be more acceptable when the
protesters' cause is conservative and system defending (against
defunding the police, against restricting abortion). In this way, we
test the hypothesis that people will judge the same protest actions as
more acceptable if the cause of the protest aligns with their own
ideological positions.

@@

In Study 1, we compiled protest actions from participants' responses and
other sources. We recruited 60 participants from the Prolific subject
pool, all of whom were citizens of the UK or the US. To increase the
socioeconomic diversity of our sample, we recruited 30 non-students
without a university degree, 15 non-students with a university degree,
and 15 current university students. Participants first read an
accessible definition of what a social group is and that collective
action is any action group members take to promote the interests of
their social group.

\begin{quote}
Society is not only made up of individuals, but consists of `social
groups' to which these individuals belong. Each person lives in a place,
has a job (or not) at an organisation, is a fan of a specific sports
club, has a religion, or belongs to any number of other such groups.
Individuals often act in ways to promote the interests of the social
groups to which they belong.
\end{quote}

\noindent Participants were then asked to name between five and ten
actions that fit that definition. Participants were encouraged to think
of actions that varied in how acceptable they were in their opinion. We
recoded responses into a smaller set of unique actions, which we
supplemented with protest actions from the psychological and political
science literature (e.g., Sharp, 1973). This process resulted in 72
actions that varied in how acceptable we would expect them to be. For
details, see Supplemental Online Material (SOM).

@@

@@

We estimated a graded response model (Bürkner, 2021; Samejima,
1997)---an item response theory model for ordinal response
variables---for participants' ratings of how often an action would be an
acceptable form of collective action.\footnote{Ratings of how often a
  given action would be an acceptable form of collective action were
  strongly and negatively correlated with how disruptive, violent, and
  extreme participants considered the action to be and how negative they
  felt about it (Table S3).} Based on the results, we selected the 25
most discriminating, informative, and relevant protest actions to form
an instrument for measuring double standards in judging collective
action in Experiments 1--3. We selected controversial, and thus
diagnostic, actions (e.g., disrupting traffic) over actions that almost
all respondents considered acceptable (e.g., signing petitions) or
unacceptable (e.g., political violence; Figure \ref{fig:f2}). For
detailed ratings and results, see SOM.

@@

In Experiment 1, we used the newly developed instrument to test for
double standards in judging collective action for workers' rights in the
United Kingdom. That is, we tested whether people with either
working-class or professional jobs applied different standards when
judging collective action taken by people with either working-class or
middle-class jobs to protest against a fictitious government bill
threatening their group's rights.

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that participants would judge the
same protest actions as more acceptable if the protesters were ingroup
members protesting for the ingroup's rights than if they were outgroup
members protesting for the outgroup's rights. By comparing reactions to
working-class and middle-class protesters, we further tested how
observers judge collective action by lower-status (i.e.~working-class)
and higher-status (middle-class) group members.\footnote{For Experiments
  1 and 2, we had preregistered the additional hypotheses that all
  participants would judge the same actions as more acceptable when the
  protesters were from the higher-status group rather than the
  lower-status group (Experiment 1) and when the goal of the protest was
  to support rather than challenge the system (Experiment 2). These
  hypotheses reflected the assumption that the motivation to justify the
  system was, to some extent, universal. With hindsight, we no longer
  think that this assumption is tenable and, therefore, omit the
  hypotheses based on it from the revised manuscript. We still report
  the same preregistered analyses and their results.} In
non-preregistered analyses, we tested whether participants would judge
the same protest actions as more or less acceptable depending on their
own ideological positions in terms of political orientation,
system-justifying beliefs, and social dominance orientation.\footnote{In
  Experiment 1, we explored a broader range of operationalizations of
  political ideology, including social dominance orientation, than in
  the other experiments.} In this way, Experiment 1 provided a first
test of the hypothesized class- and ideology-based double standards in
judging collective action.

@@

We recruited 515 participants from the Prolific subject pool who were UK
citizens, 25 years old or older, and not current students.\footnote{We
  had preregistered a sample of 500 participants, before exclusions, but
  included another 15 participants who completed the study without
  returning an approval code to the Prolific platform. Participants
  received, on average, \pounds 9.00 (\$10.93) per hour of
  participation.} As preregistered, we excluded 71 participants who
failed an attention check. This resulted in a final sample of 443
participants (\(\textit{Mdn} = 41\) years, age range: 25--76 years; 272
women, 171 men, 1 non-binary). Of these, 210 participants considered
their past, current, or future jobs to be working-class jobs.
Participants in this lower-status group did not have a university degree
and placed themselves on the bottom three ranks of the subjective
socioeconomic status ladder. Another 233 participants considered their
past, current, or future jobs to be middle-class/professional jobs.
Participants in this higher-status group had at least an undergraduate
degree and placed themselves on the top four ranks of the subjective
socioeconomic status ladder. On average, participants tended to describe
their political orientation as somewhat left-wing
(\(\textit{M} = 3.54\), \(\textit{SD} = 1.31\)) with \(36\%\) and
\(20\%\) describing their political orientation as, respectively, at
least somewhat left-wing and right-wing.

@@

Figure \ref{fig:f3} (A) shows estimates for each combination of the
protesters' and the participants' group membership. Overall, we found
that participants' responses depended on both the protesters' and the
participants' group status---but not in the directions predicted by our
hypotheses. Contradicting our hypothesis, participants did not consider
protest actions performed by their ingroup to be more acceptable, on
average, than the same actions performed by the relevant outgroup
(Cohen's \(d = 0.00, [-0.20, 0.20]\)). Instead, we found that
participants from higher-status backgrounds considered protest actions
performed by both lower-status (\(d = 0.34, [0.07, 0.59\){]}) and
higher-status (\(d = 0.33, [0.03, 0.63]\)) protesters to be, on average,
more acceptable than participants from lower-status backgrounds. All
participants considered protest actions performed by higher-status
protesters to be less acceptable, on average, than the same actions
performed by lower-status protesters (\(d = -0.20, [-0.40, -0.00]\)).

@@

\hypertarget{theoretical-implications}{%
\subsubsection{Theoretical
Implications}\label{theoretical-implications}}
